Seminar 8 DNA methylation analysis with Illumina's Infinium HumanMethylation450K array
========================================================

Ref1: http://www.ugrad.stat.ubc.ca/~stat540/seminars/seminar08_methylation.html


*Excerpt from ref*
This seminar is inspired by a STAT 540 project in 2013: Analysis of Gene Expression Omnibus Leukaemia Data from the Illumina HumanMethylation450 Array by Alice Zhu, Rachel Edgar, Shaun Jackman and Nick Fishbane. See their project website here.

```{r}

  library(GEOquery)
  library(wateRmelon)
  library(IlluminaHumanMethylation450k.db)
```


## Explore 450k methylation array data

Download/load the dataset
```{r}
  setwd("/Users/sohrab/Me/Apply/Canada Apply/Courses/Second Semester/Stat 540/Lab/stat540-2014-salehi-sohrab/data")

  if (file.exists("methyl_ALL.Rdata")) {
    # if previously downloaded
    load("methyl_ALL.Rdata")
} else {
    # if downloading for the first time
    GSE39141 <- getGEO("GSE39141")
    show(GSE39141)  ## 33 samples (29 ALL and 4 healthy B cells)
    GSE42865 <- getGEO("GSE42865")  # took ~2 mins for JB
    show(GSE42865)  ## 16 samples (9 healthy cells B cells and 7 other cells)

    # Extract expression matrices (turn into data frames at once)
    ALL.dat <- as.data.frame(exprs(GSE39141[[1]]))
    CTRL.dat <- as.data.frame(exprs(GSE42865[[1]]))

    # Obtain the meta-data for the samples and rename them perhaps?
    ALL.meta <- pData(phenoData(GSE39141[[1]]))
    CTRL.meta <- pData(phenoData(GSE42865[[1]]))

    # create some labels
    ALL.meta$Group <- c(rep("ALL", 29), rep("HBC", 4))
    ## ALL: Case; HBC: Healthy B Cells

    # Subset both meta-data and data for control (healthy) donors
    CTRL.meta <- droplevels(subset(CTRL.meta, grepl("Healthy donor", characteristics_ch1.1)))
    CTRL.dat <- subset(CTRL.dat, select = as.character(CTRL.meta$geo_accession))

    # Rename variables
    names(ALL.dat) <- paste(ALL.meta$Group, gsub("GSM", "", names(ALL.dat)), 
        sep = "_")
    names(CTRL.dat) <- paste("HBC", gsub("GSM", "", names(CTRL.dat)), sep = "_")

    # save the data to avoid future re-downloading
    save(ALL.dat, CTRL.dat, ALL.meta, CTRL.meta, file = "methyl_ALL.Rdata")
}
```
Density plot of average Beta values for probes in the two datasets:
```{r}


```


## Normalization

Using `wateRmelon`, quantile normalize the experiments via `betaqn` method. There're 15 different normalization methods. Let's see what [this](http://www.biomedcentral.com/1471-2164/14/293) paper has to say...

$β = M/(M + U + 100)$ where $M$ indicates intensity of methylated and $U$ not methylated. They conclude that quantile normalization yeilds `markedly` improved methods, using the 3 metrics they've derived.

Check out [this paper](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3012676/) for more info on $\beta$ and $M$ values.

```{r}
  beta.matrix <- as.matrix(cbind(ALL.dat, CTRL.dat))
  str(beta.matrix, max.level = 0)
  system.time(beta.norm <- betaqn(beta.matrix))
```


## M values

Assumptions of a linear model: don't have more than one mode and span $R$?
<br />
[Here's](http://people.duke.edu/~rnau/testing.htm) a reference on assumption of linear regression. The most relevant I've found is the normality of the errors. So is $\Beta$ value indicative of some kind of an error?


```{r}
  M.norm <- beta2m(beta.norm)
```

## CpG Islands

## Differential methylation analysis with limma

## Interpretation and functional enrichment analysis